### 2018-03-16 paper/5-implementations.tex
Work on S5.1-S5.2

\Make processes keys in a linear order based on a topological sort
(see~\S\ref{sec-concurrency} for parallel \Make). For each key, it builds it if
it is older than any of its dependencies. To express \Make in our framework,
we capture the persistent information it stores by a pair
\hs{(@@modTime,}~\hs{now)} comprising the \emph{file modification time} function
\hs{modTime}~\hs{::}~\hs{k}~\hs{->}~\hs{Time} and the \emph{current time} \hs{now}.
Setting aside the explicit manipulation of file modification times, which in
reality is taken care of by the file system, the \hs{make} build system defined
below captures the essence of \Make in a clear and precise manner.

\begin{minted}[xleftmargin=10pt]{haskell}
type Time = Integer
type MakeInfo k = (k -> Time, Time)
\end{minted}
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
make :: Eq k => Build Applicative (MakeInfo k) k v
make = topological process
  where
    process key deps act = do
        (modTime, now) <- getInfo <$> get
        let dirty = or [ modTime dep > modTime key | dep <- deps ]
        when dirty $ do
            v <- act
            let newModTime k = if k == key then now else modTime k
            modify $ \s -> putInfo (putValue s key v) (newModTime, now + 1)
\end{minted}

\noindent
The helper function \hs{topological}, which we describe below, calls
\hs{process} on every \hs{key} in the topological order, providing the list of
its dependencies \hs{deps} and the action \hs{act} to compute the resulting
value if it needs to be rebuilt. To determine if the \hs{key} is \hs{dirty},
\hs{process} compares its modification time with those of its dependencies. If
the \hs{key} needs to be rebuilt, the action \hs{act} is executed and the
obtained result is stored, along with an updated file modification timestamp.

The implementation of \hs{topological} encodes the dependency strategy that
\Make has chosen to use. The function itself is defined as:

\vspace{1mm}
\begin{minted}{haskell}
topological :: Eq@\,\blk{k}@ => (k -> [k] -> State@\,@(Store i k v)@\,\blk{v}@ -> State@\,@(Store i k v)@\,@(@@))
            -> Build Applicative i k v
topological process task key = execState $ forM_ chain $ \k -> do
    let fetch k = do store <- get; return (getValue store k)
    case task fetch k of
        Nothing  -> return ()
        Just act -> process k (deps k) act
  where
    deps  = dependencies task
    chain = topologicalSort deps (transitiveClosure deps key)
\end{minted}
\vspace{1mm}

\noindent
The \hs{where} clause corresponds to the pre-processing step, which uses the
function \hs{dependencies}, defined in~\S\ref{sec-deps}, to extract static
dependencies from the provided applicative \hs{task}. We compute the linear
processing \hs{chain} by taking the \hs{transitiveClosure} of \hs{key}'s
dependencies, and performing the \hs{topologicalSort} of the result. We omit
implementation of textbook graph algorithms \hs{transitiveClosure} and
\hs{topologicalSort}~\cite{cormen2001introduction}.

@@

\todo{AM}{I might do some simplification of this tomorrow. Neil: I guess it's best
for you to focus on nailing traces and associated \Shake and \Bazel sections.}

As a programming exercise, \Excel is the most challenging, due to the complicated reordering dependency scheme, which requires retrying out-of-order operations. The complete implementation is:

\begin{minted}{haskell}
excel :: Eq v => Build Monad (Changed k v, [k]) k v
excel = withChanged $ reordering $ \k dk _ act -> do
    dirty <- getChanged k ||^ maybe (return True) (anyM getChanged) dk
    if not dirty
       then return Nothing
       else act >>= \case
              Left e -> return $ Just e
              Right (_, v) -> do putStore k v
                                 return Nothing
\end{minted}

We use the helper function \hs{reordering} which provides the key to build (\hs{k}), the dependencies of \hs{k} (\hs{dk}) and an action to build \hs{k} (\hs{act}). The dirty check is the same as \hs{makeDirtyBit}, with the added complication that because the \hs{Task} is monadic it may not be possible to statically get a list of dependencies - if \hs{dk} is \hs{Nothing} then the action uses the monadic power and we conservatively rebuild anyway. If the check determines that the key is dirty, and needs rebuilding, we run \hs{act}, giving two possibilities:

\begin{enumerate}
\item The key cannot be built at this time because it depends on a key which has not yet been computed - and thus \hs{act} returns \hs{Left e}, where \hs{e} is the key that we are waiting for. In response we return \hs{e} as the result of step - asking \hs{reordering} to requeue this key after \hs{e}.
\item The key builds successfully, so we store it and return \hs{Nothing} to indicate \hs{reordering} can move on to the next key.
\end{enumerate}

The astute reader may notice that \hs{reordering} supplies extra information that is ignored with \hs{_} patterns. While not required for \Excel, we have implemented build systems using \hs{reordering} which use verifying and constructive traces, effectively turning \Excel into a cloud build system and ensuring \hs{reordering} is not overly fitted to \Excel alone (additional implementations available in the supplementary material). The definition of \hs{reordering} is:

\begin{minted}{haskell}
reordering
    :: (m ~ M (i, [k]) k v, Default i)
    => (k -> Maybe [k] -> (k -> m (Maybe v)) -> m (Either k ([k], v)) -> m (Maybe k))
    -> Build Monad (i, [k]) k v
\end{minted}

Dynamic requires a function that takes takes 4 arguments (we use \hs{m} to elide \hs{M (i, [k]) k v} from all the signatures):

\begin{enumerate}
\item \hs{k} - the key that is being built.
\item \hs{Maybe [k]} - the list of dependencies, or \hs{Nothing} if the step makes use of monadic dependencies. Note that above if a step is truly monadic it is always rebuilt. In the case of spreadsheets such steps are very rare, so a very conservative approximation can be used.\footnote{In truth, many spreadsheets take the alternative approximation of always assuming monadic steps \textit{did not} change - trading correctness for performance.}
\item \hs{k -> m (Maybe v)} - a function that looks up a key - it returns \hs{Nothing} if that key has not yet been built (and thus there is a dependency violation).
\item \hs{m (Either k ([k], v))} - an action to produce either the \hs{k} which is required but not yet computed (the reason you fail your dependency order check), or a list of the keys used and value produced.
\item It returns \hs{m (Maybe k)}, with a \hs{Nothing} to indicate the the build succeeded, or \hs{Just k} to indicate that the given key \hs{k} has not been built, and thus this key should be moved after \hs{k}.
\end{enumerate}

The implementation of \hs{dynamic} is 15 lines, but is substantially complicated by the requirement to ensure newly discovered keys are added to the list - something an actual spreadsheet with a fixed universe of cells need not do.

>>>>

We provide the complete implementation of \Make in our framework in
Fig.~\ref{fig-make-implementation}. \Make processes keys in a linear order based
on the topological sort (see~\S\ref{sec-concurrency} for parallel \Make). For
each key, it builds it if it is older than any of its dependencies. We capture
the persistent build information that \Make stores by a pair
\hs{(@@modTime,}~\hs{now)} comprising the \emph{file modification time} function
\hs{modTime}~\hs{::}~\hs{k}~\hs{->}~\hs{Time} and the \emph{current time} \hs{now}.
Setting aside the explicit manipulation of file modification times, which in
reality is taken care of by the file system, the function \hs{make} captures
the essence of \Make in a clear and precise manner.

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Persistent build information
type Time       = Integer
type MakeInfo k = (k -> Time, Time)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Make build system
make :: Eq k => Build Applicative (MakeInfo k) k v
make = topological process
  where
    process key deps act = do
        (modTime, now) <- getInfo <$> get
        let dirty = or [ modTime dep > modTime key | dep <- deps ]
        when dirty $ do
            v <- act
            let newModTime k = if k == key then now else modTime k
            modify $ \s -> putInfo (putValue s key v) (newModTime, now + 1)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard graph algorithms (implementation omitted)
transitiveClosure :: Eq a => (a -> [a]) ->  a  -> [a]
topologicalSort   :: Eq a => (a -> [a]) -> [a] -> [a]
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Topological dependency strategy
topological :: Eq k => (k -> [k] -> State (Store i k v) v -> State (Store i k v) (@@))
            -> Build Applicative i k v
topological process task key = execState $ forM_ chain $ \k -> do
    let fetch k = do store <- get; return (getValue store k)
    case task fetch k of
        Nothing  -> return ()
        Just act -> process k (deps k) act
  where
    deps  = dependencies task
    chain = topologicalSort deps (transitiveClosure deps key)
\end{minted}
\vspace{-2mm}
\caption{An implementation of \Make using our framework.}\label{fig-make-implementation}
\vspace{-2mm}
\end{figure}

The helper function \hs{topological} calls \hs{process} on every \hs{key} in
the topological order, providing the list of its dependencies \hs{deps} and the
action \hs{act} to compute the resulting value if it needs to be rebuilt. To
determine if the \hs{key} is \hs{dirty}, \hs{process} compares its modification
time with those of its dependencies. If the \hs{key} needs to be rebuilt, the
action \hs{act} is executed and the obtained result is stored, along with an
updated file modification timestamp.

The implementation of \hs{topological} encodes the dependency strategy that
\Make has chosen to use. The \hs{where} clause corresponds to the pre-processing
stage, which uses the function \hs{dependencies}, defined in~\S\ref{sec-deps},
to extract static dependencies from the provided applicative \hs{task}. We
compute the linear processing \hs{chain} by taking the \hs{transitiveClosure}
of \hs{key}'s dependencies, and performing the \hs{topologicalSort} of the
result. We omit implementation of textbook graph algorithms
\hs{transitiveClosure} and \hs{topologicalSort}, e.g.
see~\cite{cormen2001introduction}.

@@

As a programming exercise, \Excel is the most challenging, due to the
complicated reordering dependency strategy that requires retrying out-of-order
operations, see Fig.~\ref{fig-excel-implementation}.

\Excel's persistently stored information is a triple: (i) the dirty bit
function \hs{k}~\hs{->}~\hs{Bool}, (ii) an approximation of key dependencies
\hs{k}~\hs{->}~\hs{DependencyApproximation}~\hs{k} that \Excel uses to handle
dynamic dependencies, and (iii) the calculation chain \hs{[@@k]} recorded in the
previous build (\S\ref{sec-background-excel}).

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Approximation of task dependencies
data DependencyApproximation k = SubsetOf [k] | Unknown
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Persistent build information
type CalcChain k = [k]
type ExcelInfo k = ((k -> Bool, k -> DependencyApproximation k), CalcChain k)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Result of speculative task execution
data Result k v = MissingDependency k | Result v [k]
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Reordering dependency strategy (implementation omitted)
reordering :: Ord k
    => (k -> State (Store i k v) (Result k v) -> State (Store i k v) (Maybe (Result k v)))
    -> Build Monad (i, CalcChain k) k v
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Excel build system
excel :: Ord k => Build Monad (ExcelInfo k) k v
excel = reordering process
  where
    process key act = do
        (dirty, deps) <- getInfo <$> get
        let rebuild = dirty key || case deps key of SubsetOf ks -> any dirty ks
                                                    Unknown     -> True
        if not rebuild
            then return Nothing
            else do
                result <- act
                case result of
                    MissingDependency _ -> return (Just result)
                    Result v dynamicDependencies -> do
                        let newDirty k = if k == key then True else dirty k
                        modify $ \s -> putInfo (putValue s key v) (newDirty, deps)
                        return (Just result)
\end{minted}
\vspace{-2mm}
\caption{An implementation of \Excel using our framework.}\label{fig-excel-implementation}
\vspace{-2mm}
\end{figure}

The helper function \hs{reordering}, whose implementation we omit since it is
technical and not particularly enlightening, calls the function \hs{process} to
\emph{try to build} a \hs{key} by executing the action \hs{act}, in the order
determined by the calculation chain. To decide whether to \hs{rebuild} the
\hs{key}, \hs{process} checks if the \hs{key} itself is marked dirty or the
approximation of its dependencies contains a dirty key. Notice that if the
dependencies of the \hs{key} are \hs{Unknown}, e.g. when it uses the
\cmd{INDIRECT} function, the \hs{key} is always rebuilt. If a \hs{rebuild} is
not needed, \hs{process} returns \hs{Nothing} to indicate this. Otherwise, it
executes \hs{act} leading to two possible \hs{result}s:

\begin{itemize}
    \item \hs{MissingDependency}~\hs{k} indicates that the execution of \hs{act} has
    failed, because one of its dependencies \hs{k} was out-of-date, i.e. the
    calculation chain from the previous build was incorrect and therefore needs
    to be reordered, deferring the \hs{key} to be rebuilt later.
    \item \hs{Result}~\hs{v}~\hs{dynamicDependencies} indicates that the
    execution has succeeded producing the value~\hs{v} and the list of the
    \hs{key}'s dynamic dependencies. We store the value, and mark it dirty to
    trigger the rebuilding of keys that depend on it.
\end{itemize}

In both of the above cases, we notify the parent \hs{reordering} function about
the outcome by returning \hs{Just}~\hs{result}. The astute reader may notice
that \hs{process} ignores \hs{dynamicDependencies} available in the \hs{result}.
While not required for \Excel, we have implemented build systems using
\hs{reordering} which use verifying and constructive traces, effectively turning
\Excel into a cloud build system and ensuring \hs{reordering} is not overly
fitted to \Excel alone (additional implementations will be submitted to the
Artefact Evaluation).
